---
title: "Dynamic Brownian Bridge Movement Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic Brownian Bridge Movement Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The dynamic Brownian bridge movement model (dBBMM) estimates utilisation
distributions from animal tracking data by modelling movement as a
conditional random walk between observed locations. Unlike static kernel
methods, the dBBMM accounts for the sequential nature of movement data,
the time between observations, and location error — and allows the
movement process to change along the trajectory.

This vignette demonstrates the full workflow using `move2UD` with
`move2` objects.

### Background

The Brownian bridge movement model (Horne et al. 2007) assumes that an
animal's path between two observed locations follows a conditional
Brownian motion — a random walk that starts at one location and ends at
the next, with the amount of "spread" determined by a movement variance
parameter (σ²ₘ). This variance captures how much the animal deviates
from a straight-line path between fixes.

The **dynamic** extension (Kranstauber et al. 2012) allows σ²ₘ to change
along the trajectory. A sliding window moves across the track, and at
each position the method tests whether a single variance or two different
variances (split at a breakpoint) better describe the data within that
window, using the Bayesian Information Criterion (BIC) for model
selection. This means the model automatically detects changes in
movement behaviour — for example, the transition from resting (low
variance) to active movement (high variance).

## Preparing the data

```{r setup, message=FALSE}
library(move2UD)
library(move2)
library(sf)
library(terra)
```

We use the fisher tracking data bundled with `move2`. The data must be:

- A `move2` object
- In a projected coordinate system (not longitude/latitude)
- Free of empty geometries (failed GPS fixes removed)
- Sorted by time with no duplicate timestamps

```{r prepare-data}
# Load the fisher data
fishers <- mt_read(mt_example())

# Remove empty geometries (failed GPS fixes)
fishers <- fishers[!st_is_empty(fishers), ]

# Select one individual
f1 <- fishers[mt_track_id(fishers) == "F1", ]
cat("Fisher F1:", nrow(f1), "locations\n")

# Project to an azimuthal equidistant CRS centred on the data
# This ensures distances are in metres and consistent across the study area
f1_proj <- st_transform(f1, mt_aeqd_crs(f1))
```

## Step 1: Estimate the dynamic variance

The variance estimation is the core of the dBBMM. The key parameters are:

- **`location_error`**: The GPS measurement error in metres. For typical
  consumer GPS units, 10–30 m is common.
- **`window_size`**: The number of locations in the sliding window (must be
  odd). Larger windows produce smoother variance estimates but may miss
  short behavioural changes. Smaller windows are more sensitive but noisier.
- **`margin`**: The minimum number of locations on each side of a potential
  breakpoint (must be odd). Defines the minimum segment length for
  breakpoint detection.

```{r estimate-variance}
var_obj <- mt_dbbmm_variance(f1_proj,
                              location_error = 25,
                              window_size = 31,
                              margin = 11)
var_obj
```

### Inspecting the variance

The estimated variance reflects the animal's movement process at each
point in time. High variance indicates active, wide-ranging movement;
low variance indicates resting or confined movement.

```{r plot-variance, fig.cap="Dynamic Brownian motion variance over time for Fisher F1. Peaks correspond to periods of active movement; valleys indicate resting."}
timestamps <- var_obj$track_data$timestamps
variance <- mt_motion_variance(var_obj)

plot(timestamps, variance, type = "l", col = "steelblue",
     xlab = "Time", ylab = expression(sigma[m]^2),
     main = "Dynamic BM Variance — Fisher F1")
```

### Breakpoints

The method detects breakpoints where the movement process changes. These
are stored in the variance object:

```{r breakpoints}
cat("Breakpoints detected:", length(var_obj$break_list), "\n")
```

## Step 2: Compute the utilisation distribution

The UD is computed by evaluating the Brownian bridge probability density
at regular time steps along the trajectory and accumulating the density
onto a raster grid.

```{r compute-ud}
ud <- mt_dbbmm_ud(var_obj,
                    location_error = 25,
                    ext = 1.0,
                    dim_size = 150,
                    verbose = FALSE)
```

The result is a `terra::SpatRaster` where cell values sum to 1.0:

```{r plot-ud, fig.cap="dBBMM utilisation distribution for Fisher F1. Darker areas indicate higher probability of use."}
plot(ud, main = "dBBMM Utilisation Distribution — Fisher F1",
     col = hcl.colors(50, "viridis"))
```

### One-step convenience

For quick analyses, you can skip the separate variance step and pass the
`move2` object directly to `mt_dbbmm_ud()`:

```{r one-step, eval=FALSE}
ud <- mt_dbbmm_ud(f1_proj,
                    location_error = 25,
                    window_size = 31,
                    margin = 11)
```

## Working with multiple individuals

All functions accept multi-track `move2` objects. Variance is estimated
per track, and UDs are computed on a common grid for direct comparison.

```{r multi-track}
# Project all fishers
fishers_proj <- st_transform(fishers, mt_aeqd_crs(fishers))

# Estimate variance for all tracks
var_list <- mt_dbbmm_variance(fishers_proj,
                               location_error = 25,
                               window_size = 31,
                               margin = 11)
cat("Tracks processed:", length(var_list), "\n")
cat("Track names:", paste(names(var_list), collapse = ", "), "\n")
```

```{r multi-ud, fig.cap="Utilisation distributions for two fishers on a common grid."}
# Compute stacked UDs (common grid)
# Use just two tracks for speed
var_two <- var_list[c("F1", "M1")]
uds <- mt_dbbmm_ud(var_two,
                     location_error = 25,
                     ext = 1.0,
                     dim_size = 100,
                     verbose = FALSE)

par(mfrow = c(1, 2))
plot(uds[["F1"]], main = "F1", col = hcl.colors(50, "viridis"))
plot(uds[["M1"]], main = "M1", col = hcl.colors(50, "viridis"))
```

## Parallel processing

For large datasets, the sliding window computation can be parallelised:

```{r parallel, eval=FALSE}
var_obj <- mt_dbbmm_variance(f1_proj,
                              location_error = 25,
                              window_size = 71,
                              margin = 21,
                              parallel = TRUE,
                              cores = 4)
```

This uses `parallel::mclapply()` on Unix/macOS. On Windows, it falls
back to sequential processing with a message.

## Choosing parameters

### Window size and margin

The `window_size` determines the temporal scale at which behavioural
changes are detected. A good starting point is to consider the expected
duration of behavioural bouts:

- **Short bouts** (minutes to hours): smaller windows (e.g., 21–31)
- **Long bouts** (hours to days): larger windows (e.g., 51–101)

The `margin` must be at least 3 and satisfies `window_size >= 2 * margin + 1`.
A margin of about 1/3 of the window size is typical.

### Location error

The location error should reflect the actual GPS measurement accuracy,
not an arbitrary value. For most wildlife GPS units, 10–30 m is
appropriate. Setting it too low can produce artefacts; too high
oversmooths the UD.

### Raster resolution

The `dim_size` parameter controls the number of cells along the longest
dimension. Higher values give finer resolution but increase computation
time. For exploratory work, 100 is sufficient; for publication-quality
maps, 300–500 is typical.

## References

Horne, J. S., Garton, E. O., Krone, S. M., & Lewis, J. S. (2007).
Analyzing animal movements using Brownian bridges. *Ecology*, 88(9),
2354–2363.

Kranstauber, B., Kays, R., LaPoint, S. D., Wikelski, M., & Safi, K.
(2012). A dynamic Brownian bridge movement model to estimate utilization
distributions for heterogeneous animal movement. *Journal of Animal
Ecology*, 81(4), 738–746.
