---
title: "Dynamic Bivariate Gaussian Bridges"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic Bivariate Gaussian Bridges}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The dynamic bivariate Gaussian bridge (dBGB) extends the dynamic
Brownian bridge movement model (dBBMM) by decomposing the movement
variance into two components: one **parallel** to the direction of travel
and one **orthogonal** (perpendicular) to it. This directional
factorisation provides richer information about the movement process and
produces more realistic utilisation distributions — narrower along
directed segments and wider where movement is random.

This vignette demonstrates the dBGB workflow and how to interpret the
directional variance decomposition.

### Background

In the standard dBBMM, the movement variance σ²ₘ is isotropic — the
animal is assumed to deviate equally in all directions from the expected
straight-line path between fixes. In reality, animals moving in a
directed fashion (e.g., commuting, migrating) deviate mostly along the
axis of travel (varying speed) rather than perpendicular to it (varying
course). An animal foraging within a patch, by contrast, deviates
roughly equally in all directions.

The dBGB (Kranstauber, Safi & Bartumeus 2014) captures this distinction
by estimating two variance components:

- **σ²_parallel**: variance along the direction of travel, reflecting
  speed variation within a segment.
- **σ²_orthogonal**: variance perpendicular to the direction of travel,
  reflecting course deviation.

From these, a **directionality index** can be computed:

$$I_d = \frac{\sigma^2_{parallel} - \sigma^2_{orthogonal}}{\sigma^2_{parallel} + \sigma^2_{orthogonal}}$$

where:

- $I_d \approx 0$ indicates **Brownian motion** (isotropic movement)
- $I_d > 0$ indicates **directed movement** (more variation in speed
  than in course)
- $I_d < 0$ indicates **constrained movement** (more course variation
  than speed variation — e.g., movement along a narrow corridor)

## Preparing the data

```{r setup, message=FALSE}
library(move2UD)
library(move2)
library(sf)
library(terra)
```

Data preparation is identical to the dBBMM workflow:

```{r prepare-data}
fishers <- mt_read(mt_example())
fishers <- fishers[!st_is_empty(fishers), ]
f1 <- fishers[mt_track_id(fishers) == "F1", ]
f1_proj <- st_transform(f1, mt_aeqd_crs(f1))
cat("Fisher F1:", nrow(f1_proj), "locations\n")
```

## Step 1: Estimate directional variance

The dBGB variance estimation works similarly to the dBBMM but produces
two variance components. The breakpoint detection operates independently
on the parallel and orthogonal dimensions — a change in directed
behaviour (parallel) does not necessarily coincide with a change in
course deviation (orthogonal).

```{r estimate-variance}
var_obj <- mt_dbgb_variance(f1_proj,
                             location_error = 25,
                             margin = 15,
                             window_size = 31)
var_obj
```

### Extracting the variance components

The `mt_motion_variance()` accessor returns a data.frame with the
parallel and orthogonal variances:

```{r extract-variance}
mv <- mt_motion_variance(var_obj)
head(mv, 10)
```

### Plotting the variance components

```{r plot-variance, fig.cap="Parallel (blue) and orthogonal (orange) movement variance for Fisher F1. When parallel variance exceeds orthogonal, the animal is moving directionally."}
timestamps <- var_obj$track_data$timestamps

plot(timestamps, mv$para, type = "l", col = "steelblue",
     xlab = "Time", ylab = expression(sigma^2),
     main = "dBGB Variance Components — Fisher F1",
     ylim = range(c(mv$para, mv$orth), na.rm = TRUE))
lines(timestamps, mv$orth, col = "darkorange")
legend("topright", legend = c("Parallel", "Orthogonal"),
       col = c("steelblue", "darkorange"), lty = 1, bty = "n")
```

### The directionality index

The directionality index $I_d$ summarises the movement process at each
point in time:

```{r directionality, fig.cap="Directionality index over time. Values near 0 indicate Brownian motion; positive values indicate directed movement. The nocturnal activity pattern of the fisher is visible as periodic shifts."}
I_d <- (mv$para - mv$orth) / (mv$para + mv$orth)

plot(timestamps, I_d, type = "l", col = "grey30",
     xlab = "Time", ylab = expression(I[d]),
     main = "Directionality Index — Fisher F1")
abline(h = 0, lty = 2, col = "red")
```

### Comparing dBBMM and dBGB variance

The isotropic dBBMM variance is a single number that combines the
parallel and orthogonal components. We can compare:

```{r compare-variance, fig.cap="The isotropic dBBMM variance (black) compared to the parallel (blue) and orthogonal (orange) dBGB components. The dBGB reveals that the isotropic estimate is dominated by the parallel component during directed movement."}
var_dbbmm <- mt_dbbmm_variance(f1_proj, location_error = 25,
                                 window_size = 31, margin = 11)
bm_var <- mt_motion_variance(var_dbbmm)

plot(timestamps, bm_var, type = "l", col = "black", lwd = 2,
     xlab = "Time", ylab = expression(sigma^2),
     main = "Isotropic vs Directional Variance",
     ylim = range(c(bm_var, mv$para, mv$orth), na.rm = TRUE))
lines(timestamps, mv$para, col = "steelblue")
lines(timestamps, mv$orth, col = "darkorange")
legend("topright",
       legend = c("dBBMM (isotropic)", "dBGB parallel", "dBGB orthogonal"),
       col = c("black", "steelblue", "darkorange"),
       lty = 1, lwd = c(2, 1, 1), bty = "n")
```

## Step 2: Compute the utilisation distribution

The dBGB UD uses an anisotropic Gaussian kernel at each time step. The
kernel is elongated along the direction of travel when the parallel
variance exceeds the orthogonal variance, producing a narrower UD along
directed segments.

```{r compute-ud}
ud <- mt_dbgb_ud(var_obj,
                  location_error = 25,
                  ext = 2.0,
                  dim_size = 150)
```

```{r plot-ud, fig.cap="dBGB utilisation distribution for Fisher F1. The anisotropic kernel produces an elongated UD along directed movement corridors compared to the isotropic dBBMM."}
plot(ud, main = "dBGB Utilisation Distribution — Fisher F1",
     col = hcl.colors(50, "viridis"))
```

### Comparing dBBMM and dBGB UDs

The key difference between the two methods is visible in the shape of
the UD along directed movement segments. The dBGB UD is typically
narrower along commuting corridors and wider in foraging areas.

```{r compare-ud, fig.cap="Comparison of isotropic (dBBMM, left) and directional (dBGB, right) utilisation distributions."}
ud_dbbmm <- mt_dbbmm_ud(var_dbbmm, location_error = 25,
                          ext = 2.0, dim_size = 150, verbose = FALSE)

par(mfrow = c(1, 2))
plot(ud_dbbmm, main = "dBBMM (isotropic)",
     col = hcl.colors(50, "viridis"))
plot(ud, main = "dBGB (directional)",
     col = hcl.colors(50, "viridis"))
```

## When to use dBGB vs dBBMM

| Criterion | dBBMM | dBGB |
|---|---|---|
| Movement type | General-purpose | Best when directed and random movement alternate |
| Computational cost | Faster (~3–6x) | Slower (more parameters to estimate) |
| Information content | Single variance per segment | Parallel + orthogonal + directionality index |
| UD shape | Isotropic spread around path | Anisotropic: narrow along corridors, wide in patches |
| Recommended for | Home range estimation, general UD | Migration corridors, commuting routes, behavioural analysis |

For most applications, starting with the dBBMM is appropriate. The dBGB
adds value when the research question specifically concerns directional
movement or when the animal's behaviour includes clear transitions
between directed and random phases.

## Multiple individuals

The multi-track workflow is identical to the dBBMM:

```{r multi-track, eval=FALSE}
fishers_proj <- st_transform(fishers, mt_aeqd_crs(fishers))

# Variance for all tracks
var_list <- mt_dbgb_variance(fishers_proj,
                              location_error = 25,
                              margin = 15,
                              window_size = 31)

# Stacked UDs on a common grid
uds <- mt_dbgb_ud(var_list, location_error = 25)
terra::nlyr(uds)  # one layer per track
```

## References

Kranstauber, B., Safi, K., & Bartumeus, F. (2014). Bivariate Gaussian
bridges: directional factorization of diffusion in Brownian bridge
models. *Movement Ecology*, 2(1), 5.

Kranstauber, B., Kays, R., LaPoint, S. D., Wikelski, M., & Safi, K.
(2012). A dynamic Brownian bridge movement model to estimate utilization
distributions for heterogeneous animal movement. *Journal of Animal
Ecology*, 81(4), 738–746.

Horne, J. S., Garton, E. O., Krone, S. M., & Lewis, J. S. (2007).
Analyzing animal movements using Brownian bridges. *Ecology*, 88(9),
2354–2363.
